---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Preliminaries

Load the required libraries.

```{r, imports, message=FALSE, warning=FALSE}
library(llperm)
library(coin)
library(ggplot2)
library(MASS)
library(dplyr)
library(tidyr)
library(stringr)
library(gridExtra)
```

Use parallel processing (optional).

```{r}
library(foreach)
library(doParallel)
numCores <- 16 #detectCores()
registerDoParallel(numCores)
```


## Load the data set

Load the data set. We have two data frames: OTU table and person metadata.

```{r, load_data}
fn.sequence <- system.file("extdata", "VEGA_sequences.txt", package = "llperm")
fn.metadata <- system.file("extdata", "VEGA_metadata.txt", package = "llperm")

otu.table    <- read.table(fn.sequence, sep="\t", header=T, row.names=1)
otu.metadata <- read.table(fn.metadata, sep="\t", header=T, row.names=1, stringsAsFactors=T)
```

These are all the taxa:

```{r}
otu.names <- colnames(otu.table)
```

This is a data frame of abundances as raw counts

```{r}
otu.counts <- cbind(otu.table, otu.metadata)
otu.counts$DietRandom <- otu.counts$Diet[sample(1:nrow(otu.counts))] 
otu.counts$Sample <- rownames(otu.counts)
otu.counts$library_size <- rowSums(otu.table)
```


## Real data statistics

Calculate mean and variance:

```{r}
# Long form OTU table
otu.reads <- otu.counts %>% 
  select(all_of(otu.names)) %>% 
  pivot_longer(cols=c(everything()), names_to="OTU", values_to="Reads")
# Fit Negative Binomial to read counts
fit <- glm.nb(Reads ~ 1, data=otu.reads)
alpha = fit$coefficients[["(Intercept)"]]
theta = fit$theta
# Add implied Negative Binomial variance
stats <- otu.reads %>% 
  group_by(OTU) %>% 
  summarise(Mean = mean(Reads), Variance = var(Reads)) %>% 
  mutate(NB = Mean + Mean^2/theta)

```

Library size varies and data has significant overdispersion:

```{r}
df <- gather(stats, key = Type, value = Variance, c("Variance", "NB"))

# Library size looks like slightly skewed log-normal
p1 <- ggplot(otu.counts, aes(x=library_size)) + geom_histogram(bins=30) + scale_x_continuous(trans='log10')
# Compare mean and variance, highly overdispersed
colors = c("Poisson"="grey", "NegBin"="blue", "Data"="black")
p2 <- ggplot(stats, aes(x = Mean)) + 
  geom_point(aes(y = Variance, color="Data")) + 
  geom_line(aes(y = NB, color="NegBin")) + 
  geom_abline(aes(intercept = 0, slope = 1, color="Poisson")) + 
  labs(x = "Mean", y = "Variance", color = "Legend") + scale_color_manual(values = colors) +
  scale_x_log10(labels = scales::scientific) + scale_y_log10(labels = scales::scientific) + 
  theme(legend.position="top")
grid.arrange(p1, p2, ncol=2)
```

## Functions to simulate from Zero-Inflated Negative Binomial

```{r}
library(pscl)

# Given a data frame of OTU counts and columns, calculate zero-inflated negative binomial parameters
zinb_parameters <- function(otu.counts, otu.names) {
  parameters <- list()
  for (otu in otu.names) {
    if (all(otu.counts[[otu]] > 0)) {
      fit <- glm.nb(get(otu) ~ 1, data=otu.counts)
      alpha = fit$coefficients[["(Intercept)"]]
      beta = NA
      theta = fit$theta
    } else {
      fit <- zeroinfl(get(otu) ~ 1 | 1, data = otu.counts, dist = "negbin", EM = F)
      alpha = fit$coefficients$count[["(Intercept)"]]
      beta = fit$coefficients$zero[["(Intercept)"]]
      theta = fit$theta
    }
    parameters[[otu]] <- c(alpha=alpha, beta=beta, theta=theta)
  }
  parameters <- data.frame(parameters)
  return(parameters)
}
```


```{r}
# Generate simulated signal from strata defined by a combination of features, 
# each feature value makes 10% of OTUs independently differentially abundant
generate_signal <- function (strata, sample.size, otu.names, effect=log(2)) {
  # Randomly sample a data frame of features from a given strata definition
  idx <- sample(1:nrow(strata), sample.size, replace=T, prob=strata$Probability)
  df.features <- strata[idx,]
  features <- colnames(strata)[colnames(strata) != "Probability"]
  # Loop over different features that define a strata
  diff.abundant <- list()
  for (feature in features) {
    # For every feature value, randomly sample 10% of OTUs differentially abundant
    diff.abundant.value <- list()
    values <- unique(df.features[[feature]])
    for (value in values){
      diff.abundant.value[[value]] <- sample(otu.names, length(otu.names)/10)
    }
    diff.abundant[[feature]] <- diff.abundant.value
  }
  # Define effects X
  X <- array(0, c(sample.size,length(otu.names)), list(1:sample.size, otu.names))
  # Loop over every person and define their OTU effect  
  for (i in 1:sample.size) {
    person <- df.features[i,]
    for (feature in names(diff.abundant)) {
      otus <- diff.abundant[[feature]][[person[[feature]]]]
      X[i, otus] = X[i, otus] + effect
    }
  }
  # These OTUs are truly differentially abundant (affected by Group)
  diff.abundant <- unique(unlist(diff.abundant$Group))
  return(list(df.effects=X, df.features=df.features, diff.abundant=diff.abundant))
}
```

```{r}
# Given zero-inflated negative binomial parameters and sample size, create a simulated data set
zinb_simulate <- function(parameters, sample.size, strata=NULL, effect=log(2)) {
  otu.names <- colnames(parameters)
  # Whether there is strata that defines any signal, i.e. differentially abundant OTUs
  if (!is.null(strata))
      signal <- generate_signal(strata, sample.size, otu.names, effect)
  # Generate OTU counts
  otu.simulated <- list()
  for (otu in otu.names) {
    params <- parameters[[otu]]
    alpha <- params[1]
    beta <- params[2]
    theta <- params[3]
    otu.effect = if (!is.null(strata)) signal$df.effects[,otu] else 0 # This is the 'signal'
    is.zero = if (!is.na(beta)) rbinom(sample.size, size=1, prob=1/(1+exp(-beta + otu.effect))) else rep(0,sample.size)
    count.nonzero = rnbinom(sample.size, size=theta, mu=exp(alpha + otu.effect))
    otu.simulated[[otu]] = ifelse(is.zero, 0, count.nonzero)
  }
  otu.simulated <- data.frame(otu.simulated)
  # Add features if relevant
  if (!is.null(strata)) {
    otu.simulated$library_size <- rowSums(otu.simulated)
    otu.simulated <- data.frame(otu.simulated, signal$df.features)
  }
  return(list(X=otu.simulated, 
              y=if (!is.null(strata)) signal$diff.abundant else NULL))
}
```

```{r}
# Simple statistics calculation
simple_stats <- function(labels, scores, threshold=0.05, decreasing=FALSE){
  labels <- labels[order(scores, decreasing=decreasing)]
  scores <- scores[order(scores, decreasing=decreasing)]
  # Calculate AUC
  AUC <- 1 - sum(labels * cumsum(!labels)) / (sum(!labels) * sum(labels))
  # Calculate Power at a given False Discovery Rate (FDR)
  TPR <- cumsum(labels)/sum(labels)
  FPR <- cumsum(!labels)/sum(!labels)
  PowerAt <- TPR[which.min(abs(FPR - threshold))]
  # Calculate Power = TP/P and FDR = FP/N
  labels_pred <- if (decreasing) scores > threshold else scores <= threshold
  Power   <- sum(labels_pred & labels)/sum(labels)
  FPR     <- sum(labels_pred & !labels)/sum(!labels)
  results <- data.frame(AUC=AUC, PowerAt=PowerAt, Power=Power, FPR=FPR)
  return(results)
}
```

# Compare real data and simulated data

This is the distribution of real data for 9 example taxa:

```{r}
# Plot non-zero counts and proportion of non-zero counts for 9 example taxa
otu.example <- sample(otu.names, 9)
p1 <- ggplot(otu.reads %>% filter(OTU %in% otu.example), aes(x=Reads, color=OTU, fill=OTU)) +
  geom_histogram(bins=30) + facet_wrap(~OTU) + scale_x_continuous(trans='log10') + 
  theme(legend.position="top", legend.text = element_text(size=6), legend.title = element_blank())
p2 <- ggplot(otu.reads %>% filter(OTU %in% otu.example) %>% mutate(Nonzero = Reads > 0), aes(x=Nonzero, color=OTU, fill=OTU)) +
  geom_bar(stat="count") + facet_wrap(~OTU) + theme(legend.position="top", legend.text = element_text(size=6), legend.title = element_blank())
grid.arrange(p1, p2, ncol=2)
```

Fit zero-inflated Negative Binomial to the example OTUs:

```{r}
parameters <- zinb_parameters(otu.counts, otu.example)
```

Simulate data from these parameters:

```{r}
otu.simulated <- zinb_simulate(parameters, sample.size=nrow(otu.counts))$X
```

Visualize, looks pretty good:

```{r}
otu.simulated.reads <- otu.simulated %>% pivot_longer(cols=everything(), names_to="OTU", values_to="Reads")

p1 <- ggplot(otu.simulated.reads, aes(x=Reads, color=OTU, fill=OTU)) +
  geom_histogram(bins=30) + facet_wrap(~OTU) + scale_x_continuous(trans='log10') + 
  theme(legend.position="top", legend.text = element_text(size=6), legend.title = element_blank())
p2 <- ggplot(otu.simulated.reads %>% mutate(Nonzero = Reads > 0), aes(x=Nonzero, color=OTU, fill=OTU)) +
  geom_bar(stat="count") + facet_wrap(~OTU) + theme(legend.position="top", legend.text = element_text(size=6), legend.title = element_blank())
grid.arrange(p1, p2, ncol=2)
```


## Simulate data from ZINB

Find distribution parameters by fitting it to each taxa in real data

```{r}
# Fit ZIBN to real data
parameters <- zinb_parameters(otu.counts, otu.names)
```

Generate the 'signal' by defining artificial covariates and sampling from the distribution:

```{r}
# Specify a bivariate distribution
strata <- data.frame(Group = rep(c("meateater", "fisheater", "vegetarian", "vegan"), 2), 
                     Covariate=c(rep("urbanlow", 4), rep("urbanhigh", 4)),
                     Probability=c(0.20, 0.15, 0.10, 0.05, 0.05, 0.10, 0.15, 0.20))
# Create simulated data
data <- zinb_simulate(parameters, sample.size=nrow(otu.counts), strata=strata, effect=log(2))
otu.simulated <- data$X
otu.true <- data.frame(otu=otu.names, y_true=otu.names %in% data$y %>% as.integer)
```

Define models to fit

```{r}
models <- list(
  "Poisson"    = "%s ~ Group + Covariate + offset(log(library_size))",
  "Binomial"   = "cbind(%s, library_size - %s) ~ Group + Covariate ",
  "NegativeBinomial"     = "%s ~ Group + Covariate  + offset(log(library_size))",
  "BetaBinomial"    = "cbind(%s, library_size - %s) ~ Group + Covariate ", 
  "ZIPoisson"  = "%s ~ Group + Covariate + offset(log(library_size)) | Group + Covariate + offset(log(library_size))",
  "ZIBinomial" = "cbind(%s, library_size - %s) ~ Group + Covariate | Group + Covariate",
  "ZINegativeBinomial"   = "%s ~ Group + Covariate + offset(log(library_size)) | Group + Covariate + offset(log(library_size))",
  "ZIBetaBinomial"  = "cbind(%s, library_size - %s) ~ Group + Covariate | Group + Covariate"
)
```

Loop over models, calculating the log-likelihood based p-value and PRR-test based p-value

```{r}
# Loop over models
results <- list()
for (model in names(models)) {
  print(model)
  start <- Sys.time()
  str.model <- models[[model]]
  family <- get(model, mode = "function", envir = parent.frame())
  
  # Loop over OTUs in paraller
  results.model <- foreach (i=1:length(otu.names), .combine=rbind) %dopar% {
    otu <- otu.names[[i]]
    eq <- as.formula(do.call("sprintf", as.list(c(str.model, rep(otu, str_count(str.model, "%s"))))))
    fit <- prr.test.ll(eq, otu.simulated, test.var="Group", test="both", n.rep=500, family=family)
    c(otu=otu, loglik=fit$p.value.obs, llperm=fit$p.value.sim)
  }
  # Save p-values from both log likelihood and prr-test
  results.model <- as.data.frame(results.model) %>% 
    mutate(loglik = loglik %>% as.numeric, llperm = llperm %>% as.numeric) %>% 
    inner_join(otu.true, by="otu") %>% 
    pivot_longer(c(loglik, llperm), names_to="type", values_to="p_value") %>%
    mutate(y_pred = p_value <= 0.05)
  
  results[[model]] <- results.model
  # Time per model
  end <- Sys.time()
  print(end-start)
}
results <- bind_rows(results, .id = "family") 
```
Display statistics of each model:

```{r}
# Calculate statistics for each experiment and display the mean
stats <- results %>% 
  group_by(family, type) %>% 
  group_modify(~ simple_stats(.x$y_true, .x$p_value))
print.data.frame(stats)
```


## Finally

```{r}
stopImplicitCluster()
```
